---
title: "Everyday statistics in 30 minutes"
author: "Peter Geelan-Small - Stats Central, UNSW"
date: "29th July, 2021"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts]
    lib_dir: libs
    self_contained: false
    nature:
      highlightStyle: github
      countIncrementalSlides: false
      beforeInit: "macros.js"
---



```{r setup, include = F}

knitr::opts_chunk$set(echo = F, fig.align = "center", 
                      fig.asp = 1,
                      echo = F, message = F, warning = F)

```




```{r xaringan-logo, echo = F}

library(xaringanExtra)

use_logo(
  image_url = "StatsCentralLogo_rmd.png",
  width = "110px",
  height = "128px",
  position = css_position(bottom = "-2.5em", left = "1em"),
  link_url = NULL,
  exclude_class = c("title-slide", "inverse", "hide_logo")
)

```




```{r echo = F, message = F}

library(RColorBrewer)
library(wordcloud)
library(tidyverse)
library(ggpubr)
library(GGally)
library(plotly)
library(reshape2)  ## For "acast"
library(png)
library(kableExtra)
library(emmeans)

```


<style type="text/css">
.remark-slide-content {
  font-size: 28px;
  padding: 1em 1em 1em 1em;
}
</style>



# Background

Statistics in research

- May need statistics to get *information* from your data
- Information includes some indication of how system fits together and associated uncertainty
- Start with good design - fancy statistics can't usually fix holes in design
- At start of study, need a statistical plan of how to analyse results
- To consult the statistician after an experiment is finished is often merely to ask [them] to conduct a post mortem examination. [They] can perhaps say what the experiment died of. (Ronald Fisher, 1938. *Sankhya* 4: 14-17)

---

# Statistical plan

- Research question $\;\rightarrow\;$ objectives/hypotheses
- Objectives/hypotheses framed in terms of 
    - specific *outcome* and 
    - possible explanatory or predictor variables associated with that outcome
    
How do you know what statistical analysis methods to use?
- Depends on:
    - study design - independent data or not
    - type of outcome variable - continuous, binary, count


---

# Models - are there really so many?


- 1820s - linear regression based on normal distribution
- Early 1900s - Pearson's $\chi^2$ (chi-squared) test
- 1908 - $t$ test
- 1920s or so - analysis of variance (ANOVA) and analysis of covariance (ANCOVA)
- 1972 - generalised linear models - brought models based on normal, binomial, Poisson and other distributions together


---
 
# Models - are there really so many?

Model names are an accident of history


```{r}

#words_df <- read.csv("../data/stats_tests.csv", header = F)

words_df <- read.csv("../data/stats_tests.csv", header = F)

wordcloud(words = words_df[ , 1], freq = words_df[ , 2],
          colors = brewer.pal(6, "Dark2"),
          random.color = T,
          scale = c(3, 1), rot.per = 1/4)

```



---

# Outline



---

# Outline



---

# Data

---


# References




